A quantum parallel Markov chain Monte Carlo

### 2022-03-03 main.tex
more cleanup

\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\dd}{\mbox{d}}
\newcommand{\ppsi}{\boldsymbol{\psi}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\I}{\mathbf{I}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\newcommand{\llambda}{\boldsymbol{\lambda}}

Parallel MCMC techniques use multiple proposals to obtain efficiency gains over MCMC algorithms such as Metropolis-Hastings \citep{metropolis1953equation,hastings1970monte} and its progeny that use only a single proposal.  \citet{neal2003markov} first develops efficient MCMC transitions for inferring the states of hidden Markov models by proposing a `pool' of candidate states and using dynamic programming to select among them. Next, \citet{tjelmeland2004using} considers inference in the general setting and shows how to maintain detailed balance for an arbitrary number $P$ of proposals.  Consider a probability distribution $\pi(\dd\ttheta)$ defined on $\mathbb{R}^D$ that admits a probability density $\pi(\ttheta)$ with respect to the Lebesgue measure, i.e., $\pi(\dd \ttheta)=:\pi(\ttheta)\dd\ttheta$. To generate samples from the target distribution $\pi$, we craft a kernel $P(\ttheta_0,\dd \ttheta)$ that satisfies
\begin{align}\label{eq:fixes}
\pi(A) = \int \pi(\dd \ttheta_0) P(\ttheta_0,A) \, \quad a.s.
\end{align} 
for  $A \subset \mathbb{R}^D$ for which $\pi(A) > 0$.  Letting $\ttheta_0$ denote the current state of the Markov chain, \citet{tjelmeland2004using} proposes sampling from such a kernel $P(\ttheta_0,\cdot)$ by drawing $P$ proposals $\ttheta_1,\dots,\ttheta_P$ from a distribution $Q(\ttheta_0,\dd \ttheta) =: q(\ttheta_0,\ttheta)\dd \ttheta$ and selecting the next Markov chain state from among the current and proposed states with probabilities
\begin{align}\label{eq:probs}
\pi_p = \frac{\pi(\ttheta_p) \prod_{p' \neq p} q(\ttheta_p,\ttheta_{p'})}{\sum_{p''=0}^P \pi(\ttheta_{p''}) \prod_{p' \neq p''} q(\ttheta_{p''},\ttheta_{p'})} \, , \quad p=0,\dots,P \, .
\end{align}
\citet{tjelmeland2004using} shows that the kernel $P(\ttheta_0,\cdot)$ constructed in such a manner maintains detailed balance and hence satisfies Equation \eqref{eq:fixes}.  Others have since built on this work, developing parallel MCMC methods that generate or recycle multiple proposals \citep{frenkel2004speed,delmas2009does,calderhead2014general,yang2018parallelizable,luo2019multiple,schwedes2021rao,holbrook2021generating}.  
These developments demonstrate the ability of parallel MCMC methods to alleviate inferential challenges such as multimodality and to deliver performance gains over single-proposal competitors as measured by reduced autocorrelation between samples.  In the following, we focus on the specification presented in Algorithm \ref{alg:pMCMC} but note that the techniques we present may also be effective for generalizations of this basic algorithm.

\begin{algorithm}[t!]
	\caption{Parallel MCMC \citep{tjelmeland2004using}}\label{alg:pMCMC}
	\KwData{Initial Markov chain state $\ttheta^{(0)}$; total length of Markov chain $S$; total number of proposals per iteration $P$; routine for evaluating target density $\pi(\cdot)$; routines for drawing random samples from the proposal distribution $Q(\ttheta^{(s)},\cdot)$ and from a $P+1$ discrete distribution $Discrete(\ppi)$ given some probability vector $\ppi=(\pi_0,\dots,\pi_P)$.}
	\KwResult{A Markov chain $\ttheta^{(1)}, \dots, \ttheta^{(S)}$.}
	\For{$s \in\{1,\dots,S\}$}{
		%$\sigma_\pi \gets 0$\;
		$\ttheta_0 \gets \ttheta^{(s-1)}$\;
		\For{$p' \in \{1,\dots,P\}$}{
			$\ttheta_{p'} \gets Q(\ttheta^{(s-1)},\cdot)$\;
		}
		\For{$p' \in \{0,\dots,P\}$}{
			$\pi_{p'} \gets \pi(\ttheta_{p'}) \prod_{p''\neq p'}q(\ttheta_{p'},\ttheta_{p''})$\; 
		}
		$p \gets Discrete(\ppi/\ppi^T\mathbb{1})$\;
		$\ttheta^{(s)} \gets \ttheta_p$\;
	}
	\Return{$\ttheta^{(1)}, \dots, \ttheta^{(S)}$}\ .
	\vspace{0.5em}

@@

But the real promise and power of parallel MCMC comes from its natural parallelizability \citep{calderhead2014general}.  Contemporary hardware design emphasizes architectures that enable execution of multiple mathematical operations simultaneously. Parallel MCMC techniques stand to leverage technological developments that keep modern computation on track with Moore's Law, which predicts that processing power doubles every two years.  For example, the algorithm of \citet{tjelmeland2004using} generates $P$ conditionally independent proposals and then evaluates the probabilities of Equation \eqref{eq:probs}.  One may parallelize the proposal generation step using parallel pseudorandom number generators (PRNG) such as those advanced in \citet{salmon2011parallel}. The computational complexity of the target evaluations $\pi(\ttheta_p)$ is linear in the number of proposals. This presents a significant burden when $\pi(\cdot)$ is computationally expensive, e.g., in big data settings or for Bayesian inversion, but evaluation of the target density over the $P$ proposals is again a naturally parallelizable task.  Moreover, widely available machine learning software such as \textsc{TensorFlow} allows users to easily parallelize both random number generation and target evaluations on general purpose graphics processing units (GPU) \citep{lao2020tfp}. Finally, the pairwise evaluations of the proposal density $q(\cdot,\cdot)$ in Equation \eqref{eq:probs} are $\order{P^2}$, but \citet{massive} demonstrate the natural parallelizability of such pairwise operations, obtaining orders-of-magnitude speedups with contemporary GPUs.

@@

 In light of this dilemma, we propose to use the Gumbel-max trick \citep{papandreou2011perturb} to transform the generalized accept-reject step into a discrete optimization procedure.  From there, we use quantum minimization to efficiently sample from $Discrete(\ppi)$. Crucially, we get away with quantum parallel evaluation of the target $\pi(\cdot)$ over all superposed states because the Gumbel-max trick does not rely on a normalization step: each superposed state requires no knowledge of any target evaluation other than its own.

@@

\begin{algorithm}[!t]
	\caption{The Gumbel-max trick}\label{alg:gm}
	\KwData{A vector of unnormalized log-probabilities $\llambda^*=\log \ppi +\log c$, for $\ppi$ a discrete probability vector with $P+1$ elements.}
	\KwResult{A single sample $p \sim$ \emph{Discrete}$(\ppi)$ satisfying $p\in\{0,1,\dots,P\}$.}
	\For{$p' \in\{0,1,\dots,P\}$}{
		$z_{p'} \gets Gumbel(0,1)$\;
		$\alpha^*_{p'} \gets \lambda_{p'}^* + z_{p'}$\;
	}
	$p \gets \argmax_{p'=0,\dots,P}\alpha^*_{p'}$\;
	\Return{p}\ .

@@

\subsection{Symmetric proposals}

@@

Algorithm \ref{alg:qpMCMC} presents the details of QPMCMC.  The algorithm uses conventional (perhaps parallel) computing almost entirely, excluding the one line that calls the quantum minimization algorithm presented in Algorithm \ref{alg:min}.  To keep the conventional computing burden low, we recommend

This sub-routine seeks to find the minimizer of the function
\begin{align*}
f(p)= -\big(z_{p}+ \log\pi(\ttheta_{p}) \big) \quad \mbox{over} \quad p=0,1,\dots, P,
\end{align*} 
which combines the numerators of Equation \eqref{eq:probs}, the Gumbel-max trick and a trivial negation.  As discussed in Section \ref{sec:qmin}, the sub-routine requires only $\mathcal{O}(\sqrt{P})$ oracle evaluations and, therefore, only $\mathcal{O}(\sqrt{P})$ evaluations of the target density $\pi(\cdot)$ using a quantum circuit.  In theory, a quantum computer can perform the same computations as a classical computer, but the efficiency of the target evaluations would depend on a number of factors related to the structure of the density itself, the availability of fast quantum random access memory \citep{giovannetti2008quantum} and the ability of large numbers of quantum gates to act in concert with negligible noise \citep{kielpinski2002architecture,erhard2019characterizing}.

>>>>

\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Ttheta}{\boldsymbol{\Theta}}
\newcommand{\dd}{\mbox{d}}
\newcommand{\ppsi}{\boldsymbol{\psi}}
\newcommand{\U}{\mathbf{U}}
\newcommand{\I}{\mathbf{I}}
\renewcommand{\H}{\mathbf{H}}
\newcommand{\A}{\mathbf{A}}
\newcommand{\B}{\mathbf{B}}
\newcommand{\G}{\mathbf{G}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\newcommand{\llambda}{\boldsymbol{\lambda}}

Parallel MCMC techniques use multiple proposals to obtain efficiency gains over MCMC algorithms such as Metropolis-Hastings \citep{metropolis1953equation,hastings1970monte} and its progeny that use only a single proposal.  \citet{neal2003markov} first develops efficient MCMC transitions for inferring the states of hidden Markov models by proposing a `pool' of candidate states and using dynamic programming to select among them. Next, \citet{tjelmeland2004using} considers inference in the general setting and shows how to maintain detailed balance for an arbitrary number $P$ of proposals.  Consider a probability distribution $\pi(\dd\ttheta)$ defined on $\mathbb{R}^D$ that admits a probability density $\pi(\ttheta)$ with respect to the Lebesgue measure, i.e., $\pi(\dd \ttheta)=:\pi(\ttheta)\dd\ttheta$. To generate samples from the target distribution $\pi$, we craft a kernel $P(\ttheta_0,\dd \ttheta)$ that satisfies
\begin{align}\label{eq:fixes}
\pi(A) = \int \pi(\dd \ttheta_0) P(\ttheta_0,A) \, \quad a.s.
\end{align} 
for  $A \subset \mathbb{R}^D$ for which $\pi(A) > 0$.  Letting $\ttheta_0$ denote the current state of the Markov chain, \citet{tjelmeland2004using} proposes sampling from such a kernel $P(\ttheta_0,\cdot)$ by drawing $P$ proposals $\Ttheta_{-0}=(\ttheta_1,\dots,\ttheta_P)$ from a joint distribution $Q(\ttheta_0,\dd \Ttheta_{-0}) =: q(\ttheta_0, \Ttheta_{-0})\dd  \Ttheta_{-0}$ and selecting the next Markov chain state from among the current and proposed states with probabilities
\begin{align}\label{eq:probs}
\pi_p = \frac{\pi(\ttheta_p) q(\ttheta_p, \Ttheta_{-p})}{\sum_{p'=0}^P \pi(\ttheta_{p'})q(\ttheta_{p'}, \Ttheta_{-p'})} \, , \quad p=0,\dots,P \, .
\end{align}
Here, if $\Ttheta=(\ttheta_0,\dots,\ttheta_{P})$ is the $D\times (P+1)$ matrix having the current state and all $P$ proposals as columns, then $\Ttheta_{-p}$ is the $D\times P$ matrix obtained by removing the $p$th column.
\citet{tjelmeland2004using} shows that the kernel $P(\ttheta_0,\cdot)$ constructed in such a manner maintains detailed balance and hence satisfies Equation \eqref{eq:fixes}.  Others have since built on this work, developing parallel MCMC methods that generate or recycle multiple proposals \citep{frenkel2004speed,delmas2009does,calderhead2014general,yang2018parallelizable,luo2019multiple,schwedes2021rao,holbrook2021generating}.  
These developments demonstrate the ability of parallel MCMC methods to alleviate inferential challenges such as multimodality and to deliver performance gains over single-proposal competitors as measured by reduced autocorrelation between samples.  In the following, we focus on the specification presented in Algorithm \ref{alg:pMCMC} but note that the techniques we present may also be effective for generalizations of this basic algorithm.

\begin{algorithm}[t!]
	\caption{Parallel MCMC \citep{tjelmeland2004using}}\label{alg:pMCMC}
	\KwData{Initial Markov chain state $\ttheta^{(0)}$; total length of Markov chain $S$; total number of proposals per iteration $P$; routine for evaluating target density $\pi(\cdot)$; routines for drawing random samples from the proposal distribution $Q(\ttheta^{(s)},\cdot)$ and from a $P+1$ discrete distribution $Discrete(\ppi)$ given some probability vector $\ppi=(\pi_0,\dots,\pi_P)$.}
	\KwResult{A Markov chain $\ttheta^{(1)}, \dots, \ttheta^{(S)}$.}
	\For{$s \in\{1,\dots,S\}$}{
		%$\sigma_\pi \gets 0$\;
		$\ttheta_0 \gets \ttheta^{(s-1)}$\;
		$\Ttheta_{-0} \gets Q(\ttheta_0,\cdot)$\;
		\For{$p \in \{0,\dots,P\}$}{
			$\pi_{p} \gets \pi(\ttheta_{p})\, q(\ttheta_{p},\Ttheta_{-p})$\; 
		}
		$\hat{p} \gets Discrete(\ppi/\ppi^T\mathbb{1})$\;
		$\ttheta^{(s)} \gets \ttheta_{\hat{p}}$\;
	}
	\Return{$\ttheta^{(1)}, \dots, \ttheta^{(S)}$}\ .
	\vspace{0.5em}

@@

But the real promise and power of parallel MCMC comes from its natural parallelizability \citep{calderhead2014general}.  Contemporary hardware design emphasizes architectures that enable execution of multiple mathematical operations simultaneously. Parallel MCMC techniques stand to leverage technological developments that keep modern computation on track with Moore's Law, which predicts that processing power doubles every two years.  For example, the algorithm of \citet{tjelmeland2004using} generates $P$ conditionally independent proposals and then evaluates the probabilities of Equation \eqref{eq:probs}.  One may parallelize the proposal generation step using parallel pseudorandom number generators (PRNG) such as those advanced in \citet{salmon2011parallel}. The computational complexity of the target evaluations $\pi(\ttheta_p)$ is linear in the number of proposals. This presents a significant burden when $\pi(\cdot)$ is computationally expensive, e.g., in big data settings or for Bayesian inversion, but evaluation of the target density over the $P$ proposals is again a naturally parallelizable task.  Moreover, widely available machine learning software such as \textsc{TensorFlow} allows users to easily parallelize both random number generation and target evaluations on general purpose graphics processing units (GPU) \citep{lao2020tfp}. Finally, when generating independent proposals using a proposal distribution of the form $q(\ttheta_0,\Ttheta_{-0})=\prod_{p=1}^Pq(\ttheta_0,\ttheta_{p})$, the acceptance probabilities \eqref{eq:probs} require the $\order{P^2}$ evaluation of the $P+1\choose 2$ terms $q(\ttheta_{p},\ttheta_{p'})$, but \citet{massive,holbrook2021scalable} demonstrate the natural parallelizability of such pairwise operations, obtaining orders-of-magnitude speedups with contemporary GPUs.

@@

 In light of this dilemma, we propose to use the Gumbel-max trick \citep{papandreou2011perturb} to transform the generalized accept-reject step into a discrete optimization procedure (Section \ref{sec:gm}).  From there, we use quantum minimization to efficiently sample from $Discrete(\ppi)$. Crucially, we get away with quantum parallel evaluation of the target $\pi(\cdot)$ over all superposed states because the Gumbel-max trick does not rely on a normalization step: each superposed state requires no knowledge of any target evaluation other than its own.

Once one has effectively parallelized the target evaluations $\pi(\ttheta_{p})$, the new computational bottleneck is the calculations required to obtain the $P+1$ proposal density terms $q(\ttheta_{p},\Ttheta_{-p})$, for $p=0,\dots,P$.  Under an independent proposal mechanism with a proposal distribution of the form $q(\ttheta_0,\Ttheta_{-0})=\prod_{p=1}^Pq(\ttheta_0,\ttheta_{p})$, the acceptance probabilities \eqref{eq:probs} require the $\order{P^2}$ evaluation of the $P+1\choose 2$ terms $q(\ttheta_{p},\ttheta_{p'})$.  To avoid such calculations, we follow \citet{holbrook2021generating} and use symmetric proposal distributions for which $q(\ttheta_{p},\Ttheta_{-p}) = q(\ttheta_{p'},\Ttheta_{-p'})$ for $p,p'=0,\dots,P$ (Section \ref{sec:sym}).

@@

\begin{algorithm}[!t]
	\caption{The Gumbel-max trick}\label{alg:gm}
	\KwData{A vector of unnormalized log-probabilities $\llambda^*=\log \ppi +\log c$, for $\ppi$ a discrete probability vector with $P+1$ elements.}
	\KwResult{A single sample $\hat{p} \sim$ \emph{Discrete}$(\ppi)$ satisfying $\hat{p}\in\{0,1,\dots,P\}$.}
	\For{$p \in\{0,1,\dots,P\}$}{
		$z_{p} \gets Gumbel(0,1)$\;
		$\alpha^*_{p} \gets \lambda_{p}^* + z_{p}$\;
	}
	$\hat{p} \gets \argmax_{p=0,\dots,P}\alpha^*_{p}$\;
	\Return{p}\ .

@@

\subsection{Symmetric proposals}\label{sec:sym}

@@

Algorithm \ref{alg:qpMCMC} presents the details of QPMCMC.  The algorithm uses conventional (perhaps parallel) computing almost entirely, excluding the one line that calls the quantum minimization algorithm presented in Algorithm \ref{alg:min}. 
This sub-routine seeks to find the minimizer of the function
\begin{align*}
f(p)= -\big(z_{p}+ \log\pi(\ttheta_{p}) \big) \quad \mbox{over} \quad p=0,1,\dots, P,
\end{align*} 
which combines the numerators of Equation \eqref{eq:probs}, the Gumbel-max trick and a trivial negation.  As discussed in Section \ref{sec:qmin}, the sub-routine requires only $\mathcal{O}(\sqrt{P})$ oracle evaluations and, therefore, only $\mathcal{O}(\sqrt{P})$ evaluations of the target density $\pi(\cdot)$ using a quantum circuit.  In theory, a quantum computer can perform the same computations as a classical computer, but the efficiency of the target evaluations would depend on a number of factors related to the structure of the density itself, the availability of fast quantum random access memory \citep{giovannetti2008quantum} and the ability of large numbers of quantum gates to act in concert with negligible noise \citep{kielpinski2002architecture,erhard2019characterizing}.
